---
title: "GAM"
author: "Archie Benn"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(caret)
library(zoo)
library(mgcv)
library(here)

df_08 <- read_csv(here("data/main/08_sorting/df_main.csv"))

# subset at one site for testing GAMs
# highest P/Tair average
df_GF_Guy <- df_08 %>%
    filter(Site_ID == "GF-Guy")

# lowest P/Tair average
df_CA_NS3 <- df_08 %>%
    filter(Site_ID == "CA-NS3")


```

# Fitting GAMs
## GAM background
- In essence, a GAM is a GLM composed of a sum of smooth functions rather than a sum of linear contributions  
- GAMs use smooth functions of feature variables such that nonlinear relationships can be accounted for per feature  
- GLM: $g(\mu) = g(E(y)) = b_0 + b1⋅x_1 + b_2⋅x_2 + \epsilon$  
- GAM: $g(\mu) = g(E(y)) = b_0 + f(x_1) + f(x_2)...+ f(x_n) + \epsilon$ 
- where $\epsilon$ is the error

## Feature encoding
```
# space for encoding non-numeric data
```

## Testing GAM approach with a single feature
Trying out ET vs Air Temperature at DE-Hai

```{r}
# this is essentially an ordinary linear regression model, not really a GAM 
# to use smoothing power of GAM would do ET ~ s(Tair) - will do later

# highest P/Tair site
mod1_lm <- gam(ET ~ Tair, data = df_GF_Guy)
summary(mod1_lm)
```

(Air temp is highly significant to ET at this site with R2 at ~0.32)







