---
title: "GAM"
author: "Archie Benn"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())

library(mgcViz)
library(tidyverse)
library(caret)
library(zoo)
library(mgcv)
library(here)
library(wesanderson)
library(ggeffects)

set.seed(42)

df_08 <- read_csv(here("data/main/08_sorting/df_main.csv"))
```

```{r}
# subset at one site for testing GAMs
# highest P/Tair average
df_GF_Guy <- df_08 %>%
    filter(Site_ID == "GF-Guy")

# lowest P/Tair average
df_CA_NS3 <- df_08 %>%
    filter(Site_ID == "CA-NS3")

# set site ID as a factor
df_08$Site_ID <- as.factor(df_08$Site_ID)

```

# Fitting GAMs
## GAM background
- In essence, a GAM is a GLM composed of a sum of smooth functions rather than a sum of linear contributions  
- GAMs use smooth functions of feature variables such that nonlinear relationships can be accounted for per feature  
- GLM: $g(\mu) = g(E(y)) = b_0 + b1⋅x_1 + b_2⋅x_2 + \epsilon$  
- GAM: $g(\mu) = g(E(y)) = b_0 + f(x_1) + f(x_2)...+ f(x_n) + \epsilon$ 
- where $\epsilon$ is the error

- From website: Generalized additive models (GAMs) are very helpful when analyzing complicated data that displays non-linear patterns, such as time series, and spatial data, or when the connections between predictors and the response variable are difficult to describe by straightforward linear functions.

## Feature encoding
```
# space for encoding non-numeric data if needed
```

## 'GAM' approach with a single feature (linear approach)
Trying out ET vs Air Temperature at all sites

```{r}
# this is essentially an ordinary linear regression model, not really a GAM (or seen as a special case)
# to use smoothing power of GAM would do ET ~ s(Tair) - will do later

# highest P/Tair site
mod1_lm <- gam(ET ~ Tair, data = df_08)
summary(mod1_lm)
```

(Air temp is highly significant to ET at this site with R2 at ~0.45)


## GAM (nonlinear approach)
- basis functions represent the polynomial degree contained within each knot (knots as reference points) in the GAM   
- e.g for a cubic basis each knot will be represented by a cubic polynomial function 
- f(x) is then the weighted sums of these basis functions at the knots 
- the knots are pieced together by constraining the knot polynomials at each end so they match up at 1) value, 2) 1st derivative and 3) 2nd derivative, so they're linked smoothly 
- changing the nature of the basis allows great control over the $f(x)/g(\mu)$

### Independence and random effects note
- ~ 186k observations at 68 sites means that these observations are not truly independent 
- more like 68 independent clusters rather than 186k independent points  
- so will need to add site as a random effect to account for this
- could also do climate zone/cover type as random effects but these are also contained within site_ID as each site has only one of these, so can probably leave as just site_ID as an re

```{r}
# cr = cubic regression

# k = 10 (default) -> edf = 8.996 (near max), R2 = 0.522
mod1_gam_k10 <- gam(ET ~ s(Tair, bs = "cr" , k =10) + s(Site_ID, bs = "re"), data = df_08)

# k = 20 -> edf = 18.94 (near max still), R2 = 0.526
mod1_gam_k20 <- gam(ET ~ s(Tair, bs = "cr", k =20) + s(Site_ID, bs = "re"), data = df_08)

# k = 40 (default) -> edf = 36.96, R2 = 0.527
mod1_gam_k40 <- gam(ET ~ s(Tair, bs = "cr", k =40) + s(Site_ID, bs = "re"), data = df_08)

summary(mod1_lm)
summary(mod1_gam_k20)
gam.check(mod1_gam_k20)
BIC(mod1_lm, mod1_gam_k10, mod1_gam_k20, mod1_gam_k40)
anova(mod1_lm, mod1_gam_k10, mod1_gam_k20, mod1_gam_k40, test = "Chisq")
```

- GAM fits better here (more deviance explained vs. linear model (64.4% vs 44.7%))
- `edf` for s(Tair) is 8.9 at 10 knots with R2 at 0.522, so tried k = 20 and k = 40 but this barely shifted R2 and adds complexity

## plotting ET Tair GAM 
```{r}
# make new df and add predictions from each model along range of air temps in main df and 200 temps
ET_Tair_preds <- data.frame(Tair = seq(min(df_08$Tair), max(df_08$Tair), length.out = 200))
# placeholder as needs site_ID column
ET_Tair_preds$Site_ID <- df_08$Site_ID[1]
ET_Tair_preds$lm <- predict(mod1_lm, newdata = ET_Tair_preds)
ET_Tair_preds$GAM_k10 <- predict(mod1_gam_k10, newdata = ET_Tair_preds)
ET_Tair_preds$GAM_k20 <- predict(mod1_gam_k20, newdata = ET_Tair_preds)

pred_long <- ET_Tair_preds %>% pivot_longer(-c(Tair, Site_ID), names_to = "model", values_to = "fit")

ggplot(df_08, aes(x = Tair, y = ET)) +
    geom_point(alpha = 0.01) +
    geom_line(data = pred_long, aes(x = Tair, y = fit, colour = model)) +
    
    theme_minimal() +
    xlab("Air Temperature (deg C)") +
    ylab("ET (mm/day")
```





# Fitting GAMs
```{r}
# cr = cubic regression

# 1. ALL DATA MODEL, gaussian link
ET_gam <- gam(ET ~ 
                        s(Site_ID, bs = "re") +               # site as random effect
                        s(Site_age, bs = "cr", k = 20) +
                        s(Lai_500m, bs = "cr", k = 20) +
                        s(P_sum_14D, bs = "cr", k = 20) +
                        s(SW_rad, bs = "cr", k = 20) +
                        s(Tair, bs = "cr", k = 20) +     
                        s(Wspeed, bs = "cr", k = 20) +
                        s(VPD, bs = "cr", k = 20) +
                        s(Pa, bs = "cr", k = 20)              # leaving out precip as cumulative precip is used
                        , data = df_08)

# 2. NO AGE MODEL, gauss link
ET_gam_nage <- update(ET_gam, . ~ . -s(Site_age, bs = "cr", k = 20))
```
## comparing models
```{r}
anova(ET_gam, ET_gam_nage, test = "Chisq")
```
GAM with age removed is significantly worse as a model on these data

## Plotting main GAM
Using `mgcViz` package for visualising GAMs

```{r}
gv_ET_gam <- getViz(ET_gam)  # mgcViz function
gv_ET_gam_nage <- getViz(ET_gam_nage)

check.gamViz(gv_ET_gam)    # equivalent to gam.check(ET_gam)
```
- some 

### Plots
- Each of these plots represents the predictor's contribution to the response (ET)  
- So the effective contribution at each x axis is represented in the y per plot with everything else fixed  

```{r}
gam_p1 <- plot(gv_ET_gam, allTerms = T) +
    l_fitLine(linetype = 1, colour = "red3")  +
    l_ciLine(linetype = 4, colour = "blue4") +
    l_ciBar() +
    l_rug() +
    l_points(size = 2, alpha = 0.5) + 
    theme_classic()

print(gam_p1, pages = 1)
```
