---
title: "GAMs exploration"
author: "Archie Benn"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())

library(tidyverse)
library(here)
library(wesanderson)
library(gridExtra)
library(sf)
library(rnaturalearth)
library(countrycode)
library(ggrepel)

```

# GAM results exploration
- Previous script `10_GAM_testing.R` formed 272 models and calculated predictions per site for each of the 4 models formed  
- Calculated RMSE and R^2 per site so will explore these and see if there is relatedness between strong vs. poor performing sites across these models  

```{r}
df_sites <- read_csv(here("data/main/08_sorting/sites.csv"))
df_10 <- read.csv(here("data/main/10_GAM_testing/results.csv"))
```

## R2 spread

```{r}
# full spread histgram
p1_r2 <- df_10 %>%
  
  ggplot(mapping = aes(x = bam_model, y = r2)) +
  
  # violin plot 
  geom_violin(aes(fill = bam_model)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.1, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("AsteroidCity2")) +
  ggtitle("R^2 for all sites across each GAM model") +
  theme_minimal() +
  ylab("R^2") + 
  xlab("Model") +
  scale_x_discrete(limits = c("all", "no_lai", "no_age", "no_age_no_lai"))

p1_r2

```

- $R^2$ is very consistent across all of these models

## RMSE spread

```{r}
# full spread histgram
p1_rmse <- df_10 %>%
  
  ggplot(mapping = aes(x = bam_model, y = rmse)) +
  
  # violin plot 
  geom_violin(aes(fill = bam_model)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.1, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("AsteroidCity2")) +
  ggtitle("RMSE for all sites across each GAM model") +
  theme_minimal() +
  ylab("RMSE") + 
  xlab("Model") +
  scale_x_discrete(limits = c("all", "no_lai", "no_age", "no_age_no_lai"))

p1_rmse
```

- but RMSE is not, and appears to be reduced whenever age is not included, so absolute predictions are apparently worse when site age is included despite the relative shape of observed vs. fitted from $R^2$ staying constant  
- could link back to the concurvity issues with site age and site ID: the fitted smooth for site age is introducing noise/prediction instability and cannot be fitted well 
- so the site-level 'constant' of site_ID (random effect) and site age are confounding, so the contribution/smooth function of site age is poorly formed at model inference  


# Combining these R2 and RMSE values with the main DF  
- to evaluate potential similarities within sites and their GAM predictions need to attach these values to the main df which contains metadata about climate type/other met data etc.

```{r}
# pivot wider to attach to main df by site_id later
df_10_wide <- df_10 %>%
    pivot_wider(
        names_from = bam_model,
        values_from = c(r2, rmse)
    ) %>%
    rename(Site_ID = site)

# dfs to merge
dfs_list = list(df_sites, df_10_wide)

df_merged <- dfs_list %>%
    reduce(full_join, by = "Site_ID")

summary(df_merged)

```


# looking at predictions across sites 
- will look at preds per model across various aspects of the data (as in the initial data exploration Rmd)

## 1. Climate types  
### 1.1 R2
```{r}
p_r2_climates <- df_merged %>%
  
  ggplot(mapping = aes(x = Climate_zone, y = r2_no_age)) +
  
  # violin plot 
  geom_violin(aes(fill = Climate_zone)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.1, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("Darjeeling2")) +
  ggtitle("R-squared from 'No Age' GAM across Climate Zones") +
  theme_minimal() +
  ylab("R-squared") + 
  xlab("Climate Zone") 

p_r2_climates  

```

### 1.2 RMSE
```{r}
p_rmse_climates <- df_merged %>%
  
  ggplot(mapping = aes(x = Climate_zone, y = rmse_no_age)) +
  
  # violin plot 
  geom_violin(aes(fill = Climate_zone)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.1, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("Darjeeling2")) +
  ggtitle("RMSE from 'No Age' GAM across Climate Zones") +
  theme_minimal() +
  ylab("RMSE") + 
  xlab("Climate Zone") 

p_r2_climates  

```



## 2. Cover types
### 2.1 R2

```{r}
p_r2_cover <- df_merged %>%
  
  ggplot(mapping = aes(x = Cover_type, y = r2_no_age)) +
  
  # violin plot 
  geom_violin(aes(fill = Cover_type)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.1, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("Royal2")) +
  ggtitle("R-squared from 'No Age' GAM across Cover Types") +
  theme_minimal() +
  ylab("Site Age (Years)") + 
  xlab("Cover Type") 

p_r2_cover

```

### 2.2 RMSE across cover types

```{r}
p_rmse_cover <- df_merged %>%
  
  ggplot(mapping = aes(x = Cover_type, y = rmse_no_age)) +
  
  # violin plot 
  geom_violin(aes(fill = Cover_type)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.1, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("Royal2")) +
  ggtitle("RMSE from 'No Age' GAM across Cover Types") +
  theme_minimal() +
  ylab("Site Age (Years)") + 
  xlab("Cover Type") 

p_rmse_cover

```


## 3. MAPS
- will project values onto a map to see possible relations geographically 

### 3.1 R2 world map 

```{r}
# get world data
world <- ne_countries(returnclass = "sf")

# convert site long/lat for robinson projection
sites_coords_robin <- df_merged %>%
    st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
    st_transform(crs = "+proj=robin")

p_r2_world <- world %>%
    st_transform(crs = "+proj=robin") %>%
    ggplot() +
    geom_sf(fill = "grey90", colour = "grey75", linewidth = 0.2) + 
    geom_sf(data = sites_coords_robin, aes(colour = r2_no_age)) +
    scale_colour_gradient(
        low = "tomato", high = "springgreen3",
        name = "R-squared",
        breaks = c(0, 0.25, 0.5, 0.75, 1),
        labels = c("0", "0.25", "0.5", "0.75", "1")
    ) +
    theme_linedraw() +
    theme(legend.text = element_text(size = 8),
          legend.title = element_text(size = 8),
          legend.margin = margin(r = 10)) + 
    
    ggtitle("R-squared for 'No age' model plotted geographically")

p_r2_world
```

### 3.2 RMSE world map

```{r}
# get world data
world <- ne_countries(returnclass = "sf")

# convert site long/lat for robinson projection
sites_coords_robin <- df_merged %>%
    st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
    st_transform(crs = "+proj=robin")

p_rmse_world <- world %>%
    st_transform(crs = "+proj=robin") %>%
    ggplot() +
    geom_sf(fill = "grey90", colour = "grey75", linewidth = 0.2) + 
    geom_sf(data = sites_coords_robin, aes(colour = rmse_no_age)) +
    scale_colour_gradient(
        low = "tomato", high = "springgreen3",
        name = "RMSE",
        breaks = c(0, 1, 2, 3),
        labels = c("0", "1", "2", "3")
    ) +
    theme_linedraw() +
    theme(legend.text = element_text(size = 8),
          legend.title = element_text(size = 8),
          legend.margin = margin(r = 10)) + 
    
    ggtitle("RMSE for 'No age' model plotted geographically")

p_rmse_world
```


### notes: try to find an existing ET-age relationship
- look into literature for an existing ET-age relationship as besnard did with NEP and age (f(age) predictor)  
- from this, can fit this to my data and treat it as a compact 'age effect' predictor in future GAMs  
- simplifies the model and would help with concurvity issues  
- can be taken forward to be used in ML models too  
- could also try to fit a smooth for site age separately and reduce the knots to simplify it  








