---
title: "Analysis for part 1.2"
output: html_document
date: "2026-08-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())

library(tidyverse)
library(here)
library(ltc)
library(ggrepel)
library(patchwork)
library(tikzDevice)
library(ggsignif)
```

# Analysis part 1.2
- 100 repeats of 20 randomly selected training sites and impacts of predictions on the 7 medoid test sites

```{r}
df_sites <- read_csv(here("data/main/21_analysis_1.1/cleaned_sites.csv"))
df_metrics <- read_csv(here("data/main/22_analysis_1.2/metrics_1008_1353.csv"))
df_preds <- read_csv(here("data/main/22_analysis_1.2/predictions_1008_1353.csv"))
```


```{r}
df_metrics_added <- df_metrics %>%
    
    rowwise() %>%
    
    # this sets 'updown' to 'up' if r2 improves when age is added,  and 'down' if it gets worse for colouring line between paired boxplot
    mutate(updown = case_when(
    r2_no_age > r2_age ~ "down",
    r2_no_age == r2_age ~ "no_change",
    r2_no_age < r2_age ~ "up"
  )) %>%
    
    ungroup() %>%
    
    pivot_longer(cols = c(r2_age, r2_no_age), names_to = "model_used", values_to = "value") %>% 
    mutate(model_used = factor(model_used, levels = c("r2_no_age", "r2_age"))) %>%
    
    # add cluster to name
    mutate(name_cluster = paste(test_site, " (", test_site_cluster, ")", sep = ""))

```


# testing residuals of age vs no age for R-squared with shapiro test
- run through site by site to check assumptions for t test etc.

```{r}
diffs <- df_metrics %>%
    
  filter(test_site == "US-Me3") %>%
  arrange(run) %>%
  mutate(diff = r2_age - r2_no_age) %>%
  pull(diff)

shapiro.test(diffs)
```

- repeated for the 6 test sites, only IT-SRo was non-significant, so most are not normally distributed and therefore should use wilcoxon for all sites for consistency  
- "Wilcoxon test is a good alternative to the t-test when the normal distribution of the differences between paired individuals cannot be assumed"c


```{r}
# colour palette
pal = ltc(alger)

p1 <- df_metrics_added %>%
    
    filter(test_site == "DE-Lnf") %>%
    
    # just rename the row values for these to work in the boxplot
    mutate(model_used = recode(model_used, r2_age = "Mod + Age", r2_no_age = "Mod")) %>%
    
    ggplot(aes(x = model_used, y = value)) +
    
    # colour and rename key values
    scale_colour_manual(
      name = "R$^2$ Impact",
      values = c(up = pal[2], down = pal[5], no_change = pal[1]),
      labels = c(up = "Improved", down = "Worsened", no_change = "No change")
    ) +
    
    theme_minimal() + 
    ylab("R$^2$") + 
    xlab("RF model") +

    geom_point(alpha = 0.25) +
    geom_line(aes(group=run, colour=updown), linetype="11") +
    geom_boxplot(width = 0.35, outlier.shape = NA) +
    
    annotate("text", x = -Inf, y = Inf, label = "n = 100",
              hjust = -0.5, vjust = 3, size = 3.5) +
    
    # run wilcoxon signed rank on the paired data (paired by run)
    geom_signif(comparisons = list(c("Mod + Age", "Mod")), 
                
                test = "wilcox.test", test.args = list(paired = TRUE),
                map_signif_level = TRUE,
                vjust = 0.5,
                textsize = 4.25) +
    
    facet_wrap(~name_cluster)

p1
```

```{r}
# colour palette
p2 <- df_metrics_added %>%
    
    filter(test_site != "DE-Lnf") %>%
    
    # just rename the row values for these to work in the boxplot
    mutate(model_used = recode(model_used, r2_age = "Mod + Age", r2_no_age = "Mod")) %>%
    
    ggplot(aes(x = model_used, y = value)) +
    
    # colour and rename key values
    scale_colour_manual(
      name = "R$^2$ Impact",
      values = c(up = pal[2], down = pal[5], no_change = pal[1]),
      labels = c(up = "Improved", down = "Worsened", no_change = "No change")
    ) +
    
    theme_minimal() + 
    ylab("R$^2$") + 
    xlab("RF model") +
    
    geom_point(alpha = 0.25, size = 0.5) +
    geom_line(aes(group=run, colour=updown), linetype="11") +
    geom_boxplot(width = 0.35, outlier.shape = NA) +
    
    geom_signif(comparisons = list(c("Mod + Age", "Mod")), 
                test = "wilcox.test", test.args = list(paired = TRUE),
                map_signif_level = TRUE,
                
                vjust = 0.4,
                textsize = 2) +
    
    facet_wrap(~name_cluster, scales = "free_y", nrow = 1) +
    
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5)) +

    
    theme(legend.position="none")

p2
```


```{r}
# join above two plots
p3 = p1 / p2 +
    #guide_area() +
    plot_layout(guides = 'collect', 
                axes = "collect", 
                widths = c(0.4, 1.2),
                heights = c(2, 1)
    )
p3

```


```{r}
# plot as TikZ object for integration into LaTeX
options(tikzLatexPackages = c(
    getOption("tikzLatexPackages"),
    "\\usepackage{MinionPro}\n",
    "\\usepackage{MnSymbol}\n"
))

# p1
#tikz(here("diss/figures/1.2_p1.tex"), width = 6, height = 4, standAlone = TRUE)
#print(p1)
#dev.off()

# p2
#tikz(here("diss/figures/1.2_p2.tex"), width = 6, height = 2, standAlone = TRUE)
#print(p2)
#dev.off()

# p3
tikz(here("diss/figures/1.2_p3.tex"), width = 6, height = 4, standAlone = TRUE)
print(p3)
dev.off()
```



# site and cluster stuff
- making facetted bar plote by features per cluster as a visual aid to what clusters represent


```{r}
# set features
numerical_features <- c("Site_age",
                      "ET_mean",
                      "Lai_mean",
                      "SW_rad_mean",
                      "Tair_mean",
                      "Wspeed_mean",
                      "VPD_mean",
                      "P_sum_14D_mean")

# create a summary df for the ggplot
df_summary <- df_sites %>%
    
    group_by(cluster) %>%
    
    # create summery of mean and sd for each feature within the clusters
    summarise(across(all_of(numerical_features), 
                      list(mean = mean, sd = sd), 
                     
                     # name columns as (original column) + (function name e.g mean)
                      .names = "{.col}_{.fn}")) %>%
    
    pivot_longer(
        cols = -cluster,
        names_to = c("feature", ".value"),
        names_pattern = "(.+)_(mean|sd)"
    )
```

```{r}    
pal2 = ltc(lincoln)

# now plot facets
p_clusters_bar <- df_summary %>%
    
    ggplot(aes(x = factor(cluster), y = mean, fill = factor(cluster))) +
    
    geom_col() +
    geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.25) +
    
    scale_fill_manual(values = pal2) +
    
    # wrap by features
    facet_wrap(~feature, scales = "free_y") +

    theme_minimal() +
    xlab("Cluster") +
    ylab("Mean value") +
    theme(legend.position = "none")
    
p_clusters_bar

```






















