---
title: "case study"
output: html_document
date: "2026-08-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())

library(tidyverse)
library(here)
library(ltc)
library(ggrepel)
library(gridExtra)
library(tikzDevice)
library(conflicted)
library(patchwork)
```

# analysis of case study: US-Umd and DE-Lnf medoid sites  

```{r}
conflicts_prefer(dplyr::filter)
conflicts_prefer(here::here)

df_14 <- read_csv(here("data/main/14_pre_processing/df_ml_ready.csv"))
df_stats <- read_csv(here("data/main/27_case_study/stats.csv"))
df_preds <- read_csv(here("data/main/27_case_study/preds.csv"))
df_shap_test <- read_csv(here("data/main/27_case_study/test_shap.csv"))
df_shap_train <- read_csv(here("data/main/27_case_study/train_shap.csv"))
```

## 1. pred vs. observed at these two sites
c/p from end of 1.2 .Rmd script with some adjustments for this case study 

- adapting the df 

```{r}
# attach stats to preds dfs
df_preds_stats <- df_preds %>%
    
    rename("site" = "Site_ID") %>%
    
    # and add a seasonal tag based on the month (Northern hemisphere)
    mutate(
        season = case_when(
            month(Date) %in% c(12, 1, 2) ~ "Winter",
            month(Date) %in% c(3, 4, 5) ~ "Spring",
            month(Date) %in% c(6, 7, 8) ~ "Summer",
            month(Date) %in% c(9, 10, 11) ~ "Autumn"
        )
    ) %>%
    
    # and join them
    left_join(df_stats, by = c("seed", "model", "site", "shuffled"), relationship = "many-to-one")
```


```{r}
# stats df
df_preds_stats_rounded <- df_preds_stats %>%

    # r2 and rmse per site/model/shuffled combo (not seed as want averages across all seeds)
    group_by(model, site, shuffled) %>%
    summarise(mean_r2 = round(mean(r2), 2), 
              mean_rmse = round(mean(rmse), 2),
              mean_mae = round(mean(mae), 2),
              .groups = "drop")

# for re-naming facets
model_names <- list(
    "mod" = "Mod",
    "mod+age" = "Mod + Age"
)

# as_labeller()= ggplot function to convert named vector to a labeller (to rename facets here)
# redundant now as changed names on 24-8-26 but still
mod_labeller <- as_labeller(c("Mod" = "Mod", "Mod + Age" = "Mod + Age"))

```

- and now for the plotting function

```{r}
# fro sampler
set.seed(42)

# plotter function for unshuffled preds/obs
plotter_r2 <- function(testing_site){
    
    # palettes
    pal = ltc(shuggie)
    pal_p = ltc(dora)
    
    # for colouring facets later
    #strip_colours <- if (testing_site == "US-UMd") c(pal_f[5], pal_f[1]) else c(pal_f[1], pal_f[5])

    # plot
    p_r2 <- df_preds_stats %>%
    
        # use only test site and non shuffled models
    filter(site == testing_site,
           shuffled != TRUE) %>%
    
    # take a sample of 2500 rows
    slice_sample(n = 2500) %>%

    ggplot(aes(x = observed_ET, y = predicted_ET)) +
        
    # add points and colour by season for visual aid
    geom_point(alpha = 0.4,  aes(colour = season)) +
        
    # colour and rename key values
    scale_colour_manual(
      name = "Season",
      values = pal
    ) +
        
    # 1:1 line (ie. a perfect model)
    geom_abline(slope = 1, intercept = 0, colour = pal_p[3], linetype = "dashed") +
    
    theme_bw()+
    #theme(legend.position = "bottom") +
    xlab("Observed ET (mm/day)") +
    ylab("Predicted ET (mm/day)") +
    ggtitle(testing_site) +
    
    # add mean R2 and RMSE and MAE (separately to fix in-line issue)
     geom_text(data = df_preds_stats_rounded %>% filter(site == testing_site, shuffled != TRUE),
          aes(x = -Inf, y = Inf, label = paste0("$\\overline{\\mathrm{R^2}}$ = ", mean_r2)),
          hjust = -0.15, vjust = 2, size = 3.5, inherit.aes = FALSE) +
        
    geom_text(data = df_preds_stats_rounded %>% filter(site == testing_site, shuffled != TRUE),
          aes(x = -Inf, y = Inf, label = paste0("$\\overline{\\mathrm{RMSE}}$ = ", mean_rmse, " mm/day")),
          hjust = -0.1, vjust = 4, size = 3.5, inherit.aes = FALSE) +
        
    geom_text(data = df_preds_stats_rounded %>% filter(site == testing_site, shuffled != TRUE),
          aes(x = -Inf, y = Inf, label = paste0("$\\overline{\\mathrm{MAE}}$ = ", mean_mae, " mm/day")),
          hjust = -0.1, vjust = 6, size = 3.5, inherit.aes = FALSE) +
        
        facet_wrap(~model, labeller = mod_labeller)
    
    # bit hacky but gets facet to be coloured based on r2 improvement or not when adding age
    #facet_wrap2(~pred_model, labeller = mod_labeller,
     #           strip = strip_themed(background_x = elem_list_rect(fill = strip_colours)))
    
    return(p_r2)
}
```

```{r}
# generate the plots for the two focus medoid sites
p_US_UMd <- plotter_r2("US-UMd")
p_US_UMd
```

```{r}
p_DE_Lnf <- plotter_r2("DE-Lnf")
p_DE_Lnf 
```

## 2. shap stats

```{r}
# standard error function
se <- function(x) {sd(x) / sqrt(length(x))}

# colours
pal = ltc(casa_natal)

# bar plot
p1 <- df_stats %>%
    
    # group by conditions
    group_by(site, model, shuffled) %>%
    
    # stats per combination
    summarise(mean_r2 = mean(r2),
              se_r2 = se(r2),
              sd_r2 = sd(r2),
              .groups = "drop") %>%
    
    # plot bar chart with the fill as interaction between model and shuffles boolean value
    ggplot(aes(x = interaction(model, shuffled), y = mean_r2, fill = interaction(model, shuffled))) +
    
    # columns and standard error bars
    geom_col(position = "dodge", 
             width = 0.75, 
             alpha = 0.75,
             colour = "black", 
             linewidth = 0.2) +
    
    geom_errorbar(aes(ymin = mean_r2 - se_r2, ymax = mean_r2 + se_r2),
                position = position_dodge(0.9), width = 0.2) +
    
    # aesthetics
    theme_bw() +
    ylab("R$^2$") +
    scale_fill_manual(values = pal, 
                      labels = c("mod.FALSE" = "Mod",
                                  "mod+age.FALSE" = "Mod + Age",
                                  "mod+age.TRUE" = "Mod + Shuffled Age")) +
    
    # zoom y
    coord_cartesian(ylim = c(0.75, 0.9)) +
    facet_wrap(~site) +
    xlab("") +
    theme(legend.position = "top", 
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    labs(fill = "RF Model")

p1

```

```{r}
# plot as TikZ object for integration into LaTeX
options(tikzLatexPackages = c(
    getOption("tikzLatexPackages"),
    "\\usepackage{MinionPro}\n",
    "\\usepackage{MnSymbol}\n"
))

# p1
#tikz(here("diss/figures/case_study_bars.tex"), width = 6, height = 4, standAlone = TRUE)
#print(p1)
#dev.off()

```

```{r}
# bar plot
p2 <- df_stats %>%
    
    # group by conditions
    group_by(site, model, shuffled) %>%
    
    # stats per combination
    summarise(mean_rmse = mean(rmse),
              se_rmse = se(rmse),
              sd_rmse = sd(rmse),
              .groups = "drop") %>%
    
    # plot bar chart with the fill as interaction between model and shuffles boolean value
    ggplot(aes(x = interaction(model, shuffled), y = mean_rmse, fill = interaction(model, shuffled))) +
    
    # columns and standard error bars
    geom_col(position = "dodge", 
             width = 0.75, 
             alpha = 0.85,
             colour = "black", 
             linewidth = 0.2) +
    
    geom_errorbar(aes(ymin = mean_rmse - se_rmse, ymax = mean_rmse + se_rmse),
                position = position_dodge(0.9), width = 0.2) +
    
    # aesthetics
    theme_bw() +
    ylab("RMSE") +
    scale_fill_manual(values = pal, 
                      labels = c("mod.FALSE" = "Mod",
                                  "mod+age.FALSE" = "Mod + Age",
                                  "mod+age.TRUE" = "Mod + Shuffled Age")) +
    
    # zoom y
    #coord_cartesian(ylim = c(0.75, 0.9)) +
    facet_wrap(~site) +
    xlab("") +
    theme(legend.position = "top", 
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    labs(fill = "RF Model")

#p2

```

# SHAP stuff
using a function from script 17

```{r}

# quick rename for function(messy i know)
df_preds_rename <- df_preds %>%
    rename(site = Site_ID) %>%
    mutate(model = recode(model, "mod" = "Mod",
                          "mod+age" = "Mod + Age"))
df_stats2 <- df_stats %>%
    mutate(model = recode(model, "mod" = "Mod",
                          "mod+age" = "Mod + Age"))

df_shap_test2 <- df_shap_test %>%
    mutate(model = recode(model, "mod" = "Mod",
                          "mod+age" = "Mod + Age")) 
```

```{r}
preds_with_SHAP <- function(mod, site_id, data, stats_df, shap_df, shuffled_val = FALSE){
    
    # get r2 and rmse form stats_df
    stats_sub <- stats_df %>%
        filter(model == mod, site == site_id, shuffled == shuffled_val) %>%
        summarise(r2 = mean(r2), 
                  rmse = mean(rmse),
                  mae = mean(mae))
    
    # plot ET vs time per site
    pred_p <- data %>%
        
        filter(model == mod) %>%
        filter(site == site_id) %>%
        
        # keep only every fifth day for the plot (.tex file was too big so couldn't compile)
        slice(seq(1, n(), by = 5)) %>%
        ggplot(aes(x = Date)) +
        
        
        
        # fit observed and predicted to same plot
        geom_line(aes(y = observed_ET, colour = "Observed"), alpha = 0.6) +
        geom_line(aes(y = predicted_ET, colour = "Predicted"), alpha = 0.6) +
        
        ylab("Daily ET (mm)") +
        
        theme_minimal()+
        
        # legend
        scale_colour_manual(name = NULL, 
                             values = c("Observed" = pal3[1], "Predicted" = pal3[3])) +
        
        # move legend to bottom
        theme(legend.position = "bottom") 
    
    
    # create df of SHAP score means across features
    shap_site_summary <- shap_df %>%
        
        # select chosen rows
        filter(model == mod, site == site_id, shuffled == shuffled_val) %>%
        
        # leave only actual RF model features
        select(-seed, -model, -site, -shuffled, -Site_age_raw, -Date) %>%

        # apply summaries across all columns with everything()
        # set mean_dir to now represent whether direction is +ve or -ve as men is set to abs(mean) now
        summarise(across(everything(), list(mean = function(x) mean(abs(x)), mean_dir = mean, sd = sd))) %>%
        
        # pivot across all (will drop non numeric NAs later), then name feature/value (sd or mean)
        # "_(?=mean$|mean_dir$|sd$)"is a lookahead which sees what follows the underscore and then sorts 
        pivot_longer(
            cols = matches("_(mean|mean_dir|sd)$"),
            names_to = c("feature", ".value"),
            names_sep = "_(?=mean$|mean_dir$|sd$)"
        ) %>%

        # for tikz
        mutate(feature = gsub("_", "\\\\_", feature)) %>%
        
        drop_na()
    
    
    # and now the shap plot
    shap_p <- shap_site_summary %>%
        
        # re-orders features by most important
        ggplot(aes(x = reorder(feature, mean), y = mean, fill = mean_dir > 0)) +
        
        geom_col(position=position_dodge()) +
        geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2) +
        
        # aesthetics
        theme_minimal() +
        xlab("Features in Model") +
        ylab("SHAP value") +
        ggtitle("SHAP feature importance") +
        scale_fill_manual(name = NULL,
                          values = c("TRUE" = pal3[1], "FALSE" = pal3[3]), 
                          labels = c("TRUE" = "+ Impact", "FALSE" = "- Impact")) +
        
        # move legend to bottom
        theme(legend.position = "bottom") +
        
        # flip coords to fit in feature names easily
        coord_flip()
    
    # forming the arranged plot of the above two with patchwork lib
    final_p <- pred_p + shap_p + plot_layout(widths = c(1.3, 1)) +
        plot_annotation(title = paste(mod, " at ", site_id, sep = "" ))
    
    return(final_p)
}

```


```{r}
pal3 <- ltc("dora")

# US-UMd
p_no_age_umd <- preds_with_SHAP("Mod", "US-UMd", df_preds_rename, df_stats2, df_shap_test2, shuffled_val = FALSE)
p_age_umd <- preds_with_SHAP("Mod + Age", "US-UMd", df_preds_rename, df_stats2, df_shap_test2, shuffled_val = FALSE)
p_age_umd_shuffled <- preds_with_SHAP("Mod + Age", "US-UMd", df_preds_rename, df_stats2, df_shap_test2, shuffled_val = TRUE)

# DE-Lnf
p_no_age_lnf <- preds_with_SHAP("Mod", "DE-Lnf", df_preds_rename, df_stats2, df_shap_test2, shuffled_val = FALSE)
p_age_lnf <- preds_with_SHAP("Mod + Age", "DE-Lnf", df_preds_rename, df_stats2, df_shap_test2, shuffled_val = FALSE)
p_age_lnf_shuffled <- preds_with_SHAP("Mod + Age", "DE-Lnf", df_preds_rename, df_stats2, df_shap_test2, shuffled_val = TRUE)

p_no_age_lnf
p_age_lnf
p_no_age_umd
    
```

# main combination plots for two sites

```{r}
# plot as TikZ object for integration into LaTeX
options(tikzLatexPackages = c(
    getOption("tikzLatexPackages"),
    "\\usepackage{MinionPro}\n",
    "\\usepackage{MnSymbol}\n"
))

# main us-umd
tikz(here("diss/figures/umd_main.tex"), width = 6, height = 4, standAlone = TRUE)
print(p_US_UMd)
dev.off()

tikz(here("diss/figures/umd_main2.tex"), width = 6, height = 3.5, standAlone = TRUE)
print(p_age_umd)
dev.off()



# main de-lnf
tikz(here("diss/figures/lnf_main.tex"), width = 6, height = 4, standAlone = TRUE)
print(p_DE_Lnf)
dev.off()

tikz(here("diss/figures/lnf_main2.tex"), width = 6, height = 3.5, standAlone = TRUE)
print(p_age_lnf)
dev.off()
```


# SHAP time series hetamap

```{r}
shap_heatmap <- function(shap_df, pred_data, site_id, shuffled_val = FALSE){
    
    # same palette as before
    pal3 <- ltc("dora")
    
    # first, convert to long 
    shap_long <- shap_df %>%
        
        filter(model == "Mod + Age", site == site_id, shuffled == shuffled_val) %>%
        
        # get week number
        mutate(week = floor_date(Date, "week")) %>%
        
        pivot_longer(
            cols = c(sin_DOY, cos_DOY, Lai_500m, P_sum_14D, SW_rad, Tair, Wspeed, VPD, Site_age),
            names_to = "feature",
            values_to = "shap_value"
        ) %>%
        
        # aggregate values by week and take means (so less thing grids)
        group_by(week, feature) %>%
        summarise(mean_shap = mean(shap_value), .groups = "drop") %>%
        
        # for tikz
        mutate(feature = gsub("_", "\\\\_", feature)) %>%
        
        # from forcats package - used to reorder categorical levels
        # this oves site age to top of heatmap which is what i want
        # moving to end with Inf makes it at top of heatmap for some reason
        mutate(feature = fct_relevel(feature, "Site\\_age", after = Inf))

    
    # now build the heatmap
    shap_heatmap <- shap_long %>%
        
        ggplot(aes(x = week, y = feature, fill = mean_shap)) +
            
        geom_tile() +
        
        # gradient for SHAP values
        scale_fill_gradient2(
            name = "SHAP value (weekly mean)",
            low = pal3[3],
            mid = "white",
            high = pal3[1],
            midpoint = 0
        ) +
            
        theme_minimal() +
            
        ylab("Features\n(Mod + Age)") +
            
        xlab("Date") +
            
        theme(
            axis.text.y = element_text(size = 8),
            legend.position = "bottom")
        
        
    return(shap_heatmap)

}
```

## compute differences between predictions in both models and plot on time series graph

```{r}
# function to do this
difference_plotter <- function(site_id){
    
    # first form df with diffs per prediction row
    df_diff <- df_preds_rename %>%
        
        # only non shuffles
        filter(site == site_id, shuffled == FALSE) %>%
        group_by(Date, model) %>%
        summarise(predicted_ET = mean(predicted_ET), .groups = "drop") %>%
        pivot_wider(names_from = model, values_from = predicted_ET) %>%
        
        # subtract non-age from age values
        # therefore +ve = increased ET pred with age added
        # -ve values = decreased ET pred when age added
        mutate(diff = `Mod + Age` - Mod)
    
    # now plot difference
    p_diff <- df_diff %>%
        
        # thin as before
        slice(seq(1, n(), by = 5)) %>%   
        
        ggplot(aes(x = Date, y = diff)) +
        ggtitle(site_id) +
        
        geom_hline(yintercept = 0, linetype = "dashed", colour = "firebrick") +
        
        geom_line(colour = "black", alpha = 0.75) +
        theme_minimal() +
        ylab("$\\Delta$Predicted ET\nMod+Age - Mod\n(mm/day)") +
        xlab("Date") +
        theme(legend.position = "none")
    
    return(p_diff)

}

```

## combine above two plots per site

```{r}
shap_heat_umd <- shap_heatmap(df_shap_test2, df_preds_rename, "US-UMd")
p_diff_umd <- difference_plotter("US-UMd")

# final, joined plot for differences and SHAP values from adding age model
p_umd_shap_final <- p_diff_umd / shap_heat_umd +
    plot_layout(
        heights = c(1, 2),
        axes = "collect"
    )

p_umd_shap_final
```


```{r}
shap_heat_lnf <- shap_heatmap(df_shap_test2, df_preds_rename, "DE-Lnf")
p_diff_lnf <- difference_plotter("DE-Lnf")

# final, joined plot for differences and SHAP values from adding age model
p_lnf_shap_final <- p_diff_lnf / shap_heat_lnf +
    plot_layout(
        heights = c(1, 2),
        axes = "collect"
    )

p_lnf_shap_final
```

```{r}
# save both of these plots
# plot as TikZ object for integration into LaTeX
options(tikzLatexPackages = c(
    getOption("tikzLatexPackages"),
    "\\usepackage{MinionPro}\n",
    "\\usepackage{MnSymbol}\n"
))

# main us-umd
tikz(here("diss/figures/umd_shap_diff.tex"), width = 6, height = 5, standAlone = TRUE)
print(p_umd_shap_final)
dev.off()

tikz(here("diss/figures/lnf_shap_diff.tex"), width = 6, height = 5, standAlone = TRUE)
print(p_lnf_shap_final)
dev.off()

```




















## having a look at the training set ages and ET for this case study

```{r}
# training set = all minus these two sites
df_cs_training <- df_14 %>%
    
    filter(Site_ID != "US-UMd",
           Site_ID != "DE-Lnf")
```

## mean age of training
```{r}
mean_training_age <- df_cs_training %>%
    
    summarise(mean_train_age = mean(Site_age)) %>%
    
    pull(mean_train_age)

```

- mean site age is 111.38

```{r}
mean_usd_age <- df_14 %>%
    
    filter(Site_ID == "US-UMd") %>%
    
    summarise(mean_us_umd_age = mean(Site_age)) %>%
    pull(mean_us_umd_age)


mean_lnf_age <- df_14 %>%
    
    filter(Site_ID == "DE-Lnf") %>%
    
    summarise(mean_de_lnf_age = mean(Site_age)) %>%
    pull(mean_de_lnf_age)

```

## a plot of age vs et in training data

```{r}
p_age_et <- df_cs_training %>%
    
    group_by(Site_ID) %>%
    
    # summarise mean values across training data
    summarise(mean_age = mean(Site_age),
              mean_ET = mean(ET),
              mean_peak_ET = mean(ET_90D_peak)) %>%
    
    pivot_longer(cols = c(mean_ET, mean_peak_ET),
                 names_to = "metric",
                 values_to = "value") %>%
    
    ggplot(aes(x = mean_age, y = value)) +
    
    theme_bw() +
    geom_point() +
    geom_smooth(method = "lm") +
    ylab("ET (mm/day)") +
    xlab("Mean Site Age in Training Data (years)") +
    facet_wrap(~metric)


p_age_et

```


## stats for lm for each
```{r}
df_site_means <- df_cs_training %>%
  group_by(Site_ID) %>%
  summarise(
    mean_age = mean(Site_age),
    mean_ET  = mean(ET),
    mean_peak_ET = mean(ET_90D_peak)
  )

mean_et_age_lm <- lm(mean_ET ~ mean_age, data = df_site_means)
mean_et_peak_age_lm <- lm(mean_peak_ET ~ mean_age, data = df_site_means)
```

## summarising

```{r}
summary(mean_et_age_lm)
summary(mean_et_peak_age_lm)
```
 
 



