---
title: "Data Exploration"
author: "Archie Benn"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
                     
library(tidyverse)
library(sf)
library(rnaturalearth)
library(countrycode)
library(ggrepel)
library(here)
library(skimr)
library(gridExtra)
library(tikzDevice)
```

## Data Import
- This contains all 71 sites from FLUXNET2015 with ages from besnard2018, LAI from MODIS attached, and climate/forest types attached.  
- Also datasets on yearly averages/sums and per site data
- Not scaled yet (will happen in engineering script just before ML training etc. begins)

```{r, include=FALSE}
df_05 <- read_csv(here("data/main/05_filtering/filtered_main.csv"))
df_06 <- read_csv(here("data/main/06_ET_plots/adjusted_dates_data.csv"))
df_07 <- read_csv(here("data/main/07_climates/climates_data.csv"))
df_yearly <- read_csv(here("data/main/08_sorting/yearly.csv"))
df_sites <- read_csv(here("data/main/08_sorting/sites.csv"))
```

```{r}
# quick look at structures
skim(df_07)

# and a look at yearly
skim(df_yearly)

# and sites
skim(df_sites)

```

- No missing data in main dataset


## 1. Site data
- Look at the spread of climate sites and zones across the data

```{r}
p_climates <- df_sites %>%
    ggplot() +
    aes() +
    xlab("Forest Age at Site (years)") +
    ylab("Number of Sites") +
    geom_histogram(aes(x = Site_age, fill = Climate_zone))

p_climates


```



## 2. Daily data
### 2.1 Daily ET against time per site
- ET over time at each site will be checked for any anomalies and to see general trends
- Example of ET at one site over full date range:

```{r}
p_DE_Hai <- df_06 %>%
    filter(Site_ID == "DE-Hai") %>%
    ggplot(aes(x = Date, y = ET)) +
    theme_bw() +
    scale_x_date(date_breaks = "1 year", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Adjusted daily ET over time at PA-SPn") +
    geom_line(colour = "seagreen")

p_DE_Hai

```


- Example of how adjusting the date ranges in `06_ET_plots.R` helped with ET measurements:

```{r}
p <- df_05 %>%
    filter(Site_ID == "PA-SPn") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Daily ET over original time at PA-SPn") +
    geom_line(colour = "seagreen") 

p_adjust <- df_06 %>%
    filter(Site_ID == "PA-SPn") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Adjusted daily ET over time at PA-SPn") +
    geom_line(colour = "seagreen") 

grid.arrange(p, p_adjust, ncol=2)

```



## 3. Yearly data
- To view ET across site ages cleanly the yearly df will be used  













