---
title: "Data Exploration"
author: "Archie Benn"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
                     
library(tidyverse)
library(sf)
library(rnaturalearth)
library(countrycode)
library(ggrepel)
library(here)
library(skimr)
library(gridExtra)
library(wesanderson)
library(tikzDevice)
```

## Data Import
- This contains all 71 sites from FLUXNET2015 with ages from besnard2018, LAI from MODIS attached, and climate/forest types attached.  
- Also datasets on yearly averages/sums and per site data
- Not scaled yet (will happen in engineering script just before ML training etc. begins)

```{r, include=FALSE}
df_05 <- read_csv(here("data/main/05_filtering/filtered_main.csv"))
df_06 <- read_csv(here("data/main/06_ET_plots/adjusted_dates_data.csv"))
df_08 <- read_csv(here("data/main/08_sorting/df_main.csv"))
df_yearly <- read_csv(here("data/main/08_sorting/yearly.csv"))
df_sites <- read_csv(here("data/main/08_sorting/sites.csv"))
```

```{r}
# quick look at structures
skim(df_08)

# and a look at yearly
skim(df_yearly)

# and sites
skim(df_sites)

```

- No missing data in main dataset


## 1. Site data
- Look at the spread of climate sites and zones across the data

# 1.1 Climate zones and site ages
```{r}
p_climates_ages <- df_sites %>%
  
  ggplot(mapping = aes(x = Climate_zone, y = Site_age)) +
  
  # violin plot 
  geom_violin(aes(fill = Climate_zone)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.1, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("Darjeeling2")) +
  ggtitle("Represented Site Ages Across Climate Zones") +
  theme_minimal() +
  ylab("Site Age (Years)") + 
  xlab("Climate Zone") 

p_climates_ages  

#tikz(here("diss/figures/p_climates.tex"), width = 5, height = 4, standAlone = F)
#print(p_climates)
#dev.off()

```


# 1.2 Forest cover types and site ages

```{r}
p_cover <- df_sites %>%
  
  ggplot(mapping = aes(x = Site_age)) +
    
  theme_bw() +
  
  # violin plot 
  geom_histogram(aes(fill = Cover_type), binwidth = 25) + 
    
  # aesthetics
  scale_fill_manual(values = wes_palette("Darjeeling2"),
  labels = c(
    "DBF" = "Deciduous Broadleaf",
    "EBF" = "Evergreen Broadleaf",
    "ENF" = "Evergreen Needleleaf",
    "MF"  = "Mixed Forest",
    "OSH" = "Open Shrublands")) +
    labs(fill = "Cover Type") +
    ggtitle("Forest cover types across site ages") +
    ylab("Count") + 
    xlab("Site Age (years)") 

p_cover 

```

# 1.3 Climatic types and ET

```{r}
p_climates_ET <- df_sites %>%
  
  ggplot(mapping = aes(x = Climate_zone, y = ET_mean)) +
  
  # violin plot 
  geom_violin(aes(fill = Climate_zone)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.1, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
  ggtitle("Mean ET Across Climate Zones") +
  theme_minimal() +
  ylab("Mean Daily ET at Site (mm/day)") + 
  xlab("Climate Zone") 

p_climates_ET  

```

# 1.3 Cover types and ET
```{r}
p_cover_ET <- df_sites %>%
  
  ggplot(mapping = aes(x = Cover_type, y = ET_mean)) +
  
  # violin plot 
  geom_violin(aes(fill = Cover_type)) + 
  
  # hide redundant legend
  guides(fill = "none") +
  
  # add boxplot 
  geom_boxplot(width = 0.05, col = "black")+
    
  # aesthetics
  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
  ggtitle("Mean ET Across Cover Types") +
  theme_minimal() +
  ylab("Mean Daily ET at Site (mm/day)") + 
  xlab("Cover Type") 

p_cover_ET  

```

## 2. Daily data
### 2.1 Spread of ET across full dataset
- Just to get an idea of all values wihtin the full dataset currently

```{r}
# full spread histgram
p_ET_spread <- df_08 %>%

    ggplot(aes(x = ET)) +
    geom_histogram(fill = wes_palette("Darjeeling1")[2], colour = "black", binwidth = 0.5) +
    
    theme_minimal() +
    xlab("ET (mm/day)") 


# facet plot by age range
p_ET_spread_facet <- p_ET_spread +
    
    # use mutate from plot above to facet by specified age ranges
    facet_grid(~factor(Age_range, levels = c("0-10", "11-20", "21-50", "51-100", "101-150", "151-320"))) 

grid.arrange(p_ET_spread, p_ET_spread_facet,
             top = "Spread of ET values in full dataset and within forest age ranges")
```


### 2.2 Daily ET against time per site
- ET over time at each site will be checked for any anomalies and to see general trends
- Example of ET at one site over full date range:

```{r}
p_DE_Hai <- df_06 %>%
    
    filter(Site_ID == "DE-Hai") %>%
    
    ggplot(aes(x = Date, y = ET)) +
    
    # aesthetics
    theme_bw() +
    scale_x_date(date_breaks = "1 year", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Adjusted daily ET over time at DE_Hai") +
    geom_line(colour = "seagreen")

p_DE_Hai

```


- Example of how adjusting the date ranges in `06_ET_plots.R` helped with ET measurements:

```{r}
p <- df_05 %>%
    
    filter(Site_ID == "PA-SPn") %>%
    
    ggplot(aes(x = Date, y = ET)) +
    
    # aesthetics
    theme_bw() +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Daily ET over original time at PA-SPn") +
    geom_line(colour = "seagreen") 



p_adjust <- df_08 %>%
    filter(Site_ID == "PA-SPn") %>%
    
    ggplot(aes(x = Date, y = ET)) +
    
    # aesthetics 
    theme_bw() +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Adjusted daily ET over time at PA-SPn") +
    geom_line(colour = "seagreen") 

grid.arrange(p, p_adjust, ncol=2)

```

### 2.3 ET against individual variables at DE-Hai
#### ET across Air temperatures

```{r}
p_DE_Hai_Tair <- df_06 %>%
    
    filter(Site_ID == "DE-Hai") %>%
    
    ggplot(aes(x = Tair, y = ET)) +
    
    # aesthetics
    theme_bw() +
    ylab("ET (mm/day)") +
    xlab("Average Daily Temperature (deg C)") +
    ggtitle("Daily ET against air temperature at DE_Hai") +
    geom_point(colour = wes_palette("Cavalcanti1")[4]) 

p_D_Hai_Tair_facet <- p_DE_Hai_Tair +
    facet_grid(~year(Date)) +
    ggtitle("Daily ET against air temperature at DE_Hai by year")

grid.arrange(p_DE_Hai_Tair, p_D_Hai_Tair_facet)

```

#### ET across LAI 
```{r}
p_DE_Hai_LAI <- df_06 %>%
    
    filter(Site_ID == "DE-Hai") %>%
    
    ggplot(aes(x = Lai_500m, y = ET)) +
    
    # aesthetics
    theme_bw() +
    ylab("ET (mm/day)") +
    xlab("LAI (m2/m2)") +
    ggtitle("Daily ET against LAI at DE-Hai") +
    geom_point(colour = wes_palette("Darjeeling1")[2])

p_D_Hai_LAI_facet <- p_DE_Hai_LAI +
    facet_grid(~year(Date)) +
    ggtitle("Daily ET against LAI at DE_Hai by year")

grid.arrange(p_DE_Hai_LAI, p_D_Hai_LAI_facet)
```

#### ET across Short Wave radiation
```{r}
p_DE_Hai_SWrad <- df_06 %>%
    
    filter(Site_ID == "DE-Hai") %>%
    
    ggplot(aes(x = SW_rad, y = ET)) +
    
    # aesthetics
    theme_bw() +
    ylab("ET (mm/day)") +
    xlab("SW radiation (W m-2)") +
    ggtitle("Daily ET against SW radiation at DE-Hai") +
    geom_point(colour = wes_palette("Chevalier1")[2])

p_D_Hai_SWrad_facet <- p_DE_Hai_SWrad +
    facet_grid(~year(Date)) +
    ggtitle("Daily ET against SW radiation at DE_Hai by year") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0, hjust=0)) 

grid.arrange(p_DE_Hai_SWrad, p_D_Hai_SWrad_facet)
```

Strong linear relationship with SW radiation and ET at this site.

## 3. Yearly data
- To view ET across site ages cleanly the yearly df will be used  

### 3.1 ET vs continuous age
```{r}
# ET vs age
p_ET_age <- df_yearly %>%
    
    ggplot(aes(x = Site_age, y = ET_year)) +
    
    geom_point(colour = wes_palette("Cavalcanti1")[4]) +

    
    # aesthetics
    theme_bw() +
    ylab("ET (mm/year)") +
    xlab("Forest Site Age (years)") +
    ggtitle("Yearly ET against Forest Age") +
    facet_grid(~Climate_zone)

p_ET_age

```

### 3.2 ET against age ranges within climates
```{r}
p_ET_age_ranges <- df_yearly %>%
    
    mutate(Age_range = factor(Age_range,
            levels = c("0-10", "11-20", "21-50", "51-100", "101-150", "151-320"))) %>%
    
    ggplot(aes(x = Age_range, y = ET_year)) +
    
    geom_violin(aes(fill = Age_range), levels = c("0-10", "11-20", "21-50", "51-100", "101-150", "151-320")) +
    
    # red dashed line at mean ET
    #geom_hline(yintercept = mean(df_yearly$ET_year), linetype = "dashed", col = "red") +
    
    # add boxplot 
    geom_boxplot(width = 0.1, col = "black")+
    
    # hide redundant legend
    guides(fill = "none") +
    
    # aesthetics
    scale_fill_manual(values = c(wes_palette("AsteroidCity1"), wes_palette("Cavalcanti1")[4])) +
    theme_bw() +
    xlab("Forest Age (years)") +
    ylab("ET (mm/year)")

p_ET_age_ranges
```
### 3.3 ET, annual precip, and average temperature
```{r}
p_ET_P_T <- df_yearly %>%
    ggplot(aes(x = log(P_year), y = ET_year, colour = Tair_mean)) +
    geom_point() +
    
    # aesthetics
    theme_bw() +
    ylab("Total ET (mm/year)") +
    xlab("log(Total Precipitation (mm/year))") +
    ggtitle("Annual ET against Annual Precipitation") +
    scale_color_gradient(low = "blue", high = "red") +
    labs(colour = "Mean air\nTemperature (deg C)")

p_ET_P_T
    

```


### 3.4 ET vs LAI and Air Temperature
```{r}
p_ET_LAI <- df_yearly %>%
    
    ggplot(aes(x = Lai_mean, y = ET_year, colour = Climate_zone)) +
    geom_point() +
    
    # aesthetics 
    theme_bw() +
    
    # legend aesthetics
    theme(legend.position = c(0.225, 0.875),
        legend.background = element_rect(fill = "white", color = "black"),
        legend.key.height = unit(3, 'mm')) +
    
    ylab("ET (mm/year)") +
    xlab("Mean LAI (m2/m2)") +
    ggtitle("Yearly ET against mean LAI") 


# ET vs Tair
p_ET_Tair <- df_yearly %>%
    
    ggplot(aes(x = Tair_mean, y = ET_year, colour = Climate_zone)) +
    geom_point() +
    
    # aesthetics 
    theme_bw() +
    
    # legend aesthetics
    theme(legend.position = c(0.225, 0.875),
        legend.background = element_rect(fill = "white", color = "black"),
        legend.key.height = unit(3, 'mm')) +
    
    ylab("ET (mm/year)") +
    xlab("Average Air Temperature (deg C)") +
    ggtitle("Yearly ET against yearly LAI")

grid.arrange(p_ET_LAI, p_ET_Tair, ncol=2)

```










