---
title: "Data Exploration"
author: "Archie Benn"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(rnaturalearth)
library(countrycode)
library(ggrepel)
library(here)
library(skimr)
library(gridExtra)
```

## Data Import
- Full data set from the end of `05_filtering.R`  
- This contains all 73 sites from FLUXNET2015 with ages from besnard2018 and LAI from MODIS attached  
- Not scaled yet (will happen in `06_engineer.R`, just before ML training etc. begins)

```{r}
df_06 <- read_csv(here("data/main/05_filtering/filtered_main.csv"))

# quick look at structures
skim(df_06)
```

- No missing data in main dataset

## 1. Daily ET against time per site
- ET over time at each site will be checked for any anomalies and to see general trends

```{r}
p <- df_06 %>%
    filter(Site_ID == "AU-Cum") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Daily ET over time at AU-Cum") +
    geom_line(colour = "seagreen") 
    
p
```

- Can see a potential issue with using the values in the first 3/4 of 2012 at this site  
- Will repeat for every site using a function applied to the list of site names and assess the plots afterwards  

```{r}
get_ET_sites <- function(site, data, path_to_dir){
    
    # plot ET vs time per site
    p <- data %>%
        filter(Site_ID == site) %>%
        ggplot(aes(x = Date, y = ET)) +
        geom_line(colour = "seagreen")
    
    png(here(path_to_dir, paste0(site, ".png")), width = 8, height = 4, units = "in", res = 300)
    print(p)
    dev.off()
}

# retrieve updated site IDs after some filtering
site_ids <- unique(df_06$Site_ID)

# apply function to each site                         
# walk(site_ids, \(site) get_ET_sites(site, df, "data/fluxnet/06_exploration/06.1_sites_ET_time/"))
```


### Adjusting the plots
- Viewed the plots and found a few more instances of issues of ET over the date ranges at some sites  
- When comparing with LE the pattern is the same, so while ET is dependent on LE and Tair it appears to be that LE is th e issue  
- These date ranges will be removed based on clear issues within plots (see above example for AU-Cum ET)  

```
AU-Cum: remove before November 2012 
BE-Bra: remove before January 2005 (snip)
BR-Sa1: remove 2006-2009
BR-Sa3: remove after 2004
CA-Gro: remove before 2004 and after 2014
CA-NS1: remove after September 2005
CA-NS2: remove November 2004 - May 2005 and after September 2005
CA-NS3: remove after September 2005
CA-NS5: remove after September 2005
CA-NS6: remove after September 2005
CA-NS7: remove after September 2005
CA-Qfo: remove before July 2003
CH-Dav: remove 2005-2006
CH-Lae: remove before July 2004
DE-Hai: remove after 2010
DE-Lnf: remove 2007-2010
IT-Col: remove before 2007
IT-Cpz: remove after 2009
IT-La2: REMOVE SITE
IT-Ren: remove before 2005
IT-Ro1: REMOVE SITE (check as error somewhere)
IT-Ro2: remove 2009-2010
PA-SPn: remove before March 2007 and after July 2009
US-Blo: remove after October 2007
US-GBT: remove after 2005
US-GLE: remove before 2005
US-Ha1: remove July 2009 - January 2011
US-Me3: remove after September 2009
US-Me6: remove before March 2010
US-Prr: remove before 2011
US:Syv: remove June 2007 - June 2012
US-UMd: remove before July 2007
US-WCr: remove September 2006 - January 2011
ZM-Mon: remove October 2007 and after June 2009
```

- these will be adjusted in `07_filtering2.R` and re-imported here:


```{r}
# post running 07_filtering2.R
df_06_adjusted <- read_csv(here("data/main/07_cleaning/adjusted_dates_data.csv"))

# retrieve updated site IDs after some filtering
site_ids <- unique(df_06_adjusted$Site_ID)

# re-run and save pngs to separate folder
# walk(site_ids, \(site) get_ET_sites(site, df_adjusted, "data/fluxnet/06_exploration/06.2_sites_ET_time_adjusted/"))

```

- After checking these plots individually I am happy with the coverage of ET across the sites now  
- For example, to compare one site before/after adjusting date ranges:

```{r}
p <- df_06 %>%
    filter(Site_ID == "PA-SPn") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Daily ET over original time at PA-SPn") +
    geom_line(colour = "seagreen") 

p_adjust <- df_06_adjusted %>%
    filter(Site_ID == "PA-SPn") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Daily ET over adjusted time at PA-SPn") +
    geom_line(colour = "seagreen") 

grid.arrange(p, p_adjust, ncol=2)

```



## 2. Yearly ET
- To view ET across site ages cleanly the yearly ET total will be calculated 

```{r}



```

















