---
title: "Data Exploration"
author: "Archie Benn"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
rm(list = ls())

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(rnaturalearth)
library(countrycode)
library(ggrepel)
library(here)
library(skimr)
```

## Data Import
- Full data set from the end of `05_filtering.R`  
- This contains all 73 sites from FLUXNET2015 with ages from besnard2018 and LAI from MODIS attached  
- Not scaled yet (will happen in `06_engineer.R`, just before ML training etc. begins)

```{r}
df <- read_csv(here("data/fluxnet/05_filtering/filtered_main.csv"))

# quick look at structures
skim(df)
```

- No missing data in main dataset

## 1. ET against time per site
- will assess ET over time at each site

```{r}
p <- df %>%
    filter(Site_ID == "US-UMd") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    geom_line(colour = "seagreen") 
    

p
```

- Can see a potential issue with using the values in the first 3/4 of 2012 at this site  
- Will repeat for every site using a function applied to the list of site names and assess the plots afterwards  

```{r}
get_ET_sites <- function(site, data){
    
    # plot ET vs time per site
    p <- data %>%
        filter(Site_ID == site) %>%
        ggplot(aes(x = Date, y = ET)) +
        geom_line(colour = "seagreen")
    
    png(here("data/fluxnet/06_exploration/sites_ET_time/", paste0(site, ".png")), width = 8, height = 4, units = "in", res = 300)
    print(p)
    dev.off()
}

# read site list
site_list <- read_lines(here("data/fluxnet/01_site_selection/site_list.txt"))

# apply function to each site                         
walk(site_list, \(site) get_ET_sites(site, df))
```


### Adjustingthe plots
- Viewed the plots and found a few more instances of issues of ET over the date ranges at some sites  
- When comparing with LE the pattern is the same, so while ET is dependent on LE and Tair it appears to be that LE is th e issue  
- These date ranges will be removed based on clear issues within plots (see above example for AU-Cum ET)  

```
AU-Cum: remove before November 2012 
BE-Bra: remove before January 2005 (snip)
BR-Sa1: remove 2006-2009
BR-Sa3: remove after 2004
CA-Gro: remove before 2004 and after 2014
CA-NS1: remove after September 2005
CA-NS2: remove November 2004 - May 2005 and after September 2005
CA-NS3: remove after September 2005
CA-NS5: remove after September 2005
CA-NS6: remove after September 2005
CA-NS7: remove after September 2005
CA-Qfo: remove before July 2003
CH-Dav: remove 2005-2006
CH-Lae: remove before July 2004
DE-Hai: remove after 2010
DE-Lnf: remove 2007-2010
IT-Col: remove before 2007
IT-Cpz: remove after 2009
IT-La2: REMOVE SITE
IT-Ren: remove before 2005
IT-Ro1: REMOVE SITE (check as error somewhere)
IT-Ro2: remove 2009-2010
PA-SPn: remove before March 2007 and after July 2009
US-Blo: remove after October 2007
US-GBT: remove after 2005
US-GLE: remove before 2005
US-Ha1: remove July 2009 - January 2011
US-Me3: remove after September 2009
US-Me6: remove before March 2010
US-Prr: remove before 2011
US:Syv: remove June 2007 - June 2012
US-UMd: remove before July 2007
US-WCr: remove September 2006 - January 2011
ZM-Mon: remove October 2007 and after June 2009
```



























