---
title: "Data Exploration"
author: "Archie Benn"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
                     
library(tidyverse)
library(sf)
library(rnaturalearth)
library(countrycode)
library(ggrepel)
library(here)
library(skimr)
library(gridExtra)
```

## Data Import
- This contains all 71 sites from FLUXNET2015 with ages from besnard2018, LAI from MODIS attached, and climate/forest types attached.
- Not scaled yet (will happen in engineering script just before ML training etc. begins)

```{r}
df_06 <- read_csv(here("data/main/05_filtering/filtered_main.csv"))

# quick look at structures
skim(df_06)
```

- No missing data in main dataset

## 1. Daily ET against time per site
- ET over time at each site will be checked for any anomalies and to see general trends

```{r}
p <- df_06 %>%
    filter(Site_ID == "AU-Cum") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Daily ET over time at AU-Cum") +
    geom_line(colour = "seagreen") 
    
p
```

- Can see a potential issue with using the values in the first 3/4 of 2012 at this site  
- Will repeat for every site using a function applied to the list of site names and assess the plots afterwards  

```{r}
get_ET_sites <- function(site, data, path_to_dir){
    
    # plot ET vs time per site
    p <- data %>%
        filter(Site_ID == site) %>%
        ggplot(aes(x = Date, y = ET)) +
        geom_line(colour = "seagreen")
    
    png(here(path_to_dir, paste0(site, ".png")), width = 8, height = 4, units = "in", res = 300)
    print(p)
    dev.off()
}

# retrieve updated site IDs after some filtering
site_ids <- unique(df_06$Site_ID)

# apply function to each site                         
walk(site_ids, \(site) get_ET_sites(site, df, "data/fluxnet/06_exploration/06.1_sites_ET_time/"))
```


- After checking these plots individually I am happy with the coverage of ET across the sites now  
- For example, to compare one site before/after adjusting date ranges:

```{r}
p <- df_06 %>%
    filter(Site_ID == "PA-SPn") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Daily ET over original time at PA-SPn") +
    geom_line(colour = "seagreen") 

p_adjust <- df_06_adjusted %>%
    filter(Site_ID == "PA-SPn") %>%
    ggplot(aes(x = Date, y = ET)) +
    scale_x_date(date_breaks = "3 month", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ylab("ET (mm/day)") +
    ggtitle("Adjusted daily ET over time at PA-SPn") +
    geom_line(colour = "seagreen") 

grid.arrange(p, p_adjust, ncol=2)

```



## 2. Yearly ET
- To view ET across site ages cleanly the yearly ET total will be calculated 

```{r}
# set up yearly df
df_years <- df_06_adjusted %>%
    group_by(Site_ID, year(Date)) %>%
    summarise(Site_ID = first(Site_ID),
              Latitude = first(Latitude),
              Longitude = first(Longitude),
              Cover_type = first(Cover_type),
              Site_age = first(Site_age),
              ET_year = sum(ET), 
              Lai_mean = mean(Lai_500m),
              SW_rad_mean = mean(SW_rad),
              Tair_mean = mean(Tair),
              Wspeed_mean = mean(Wspeed),
              VPD_mean = mean(VPD),
              P_year = sum(P),
              Pa_mean = mean(Pa))
              

    
p_test <- df_years%>%
    ggplot(aes(x = Site_age, y = ET_year)) +
    theme_minimal() +
    ylab("ET (mm/year)") +
    xlab("Site Age (years)") +
    ggtitle("Yearly ET against forest age at all sites") +
    geom_point() 

p_test

```






## 3. Site data
- Look at the spread of climate sites and zones across the data

### Koeggen biomes

```{r}




```










